---
title: "p8105_hw3_CY2772"
author: "Chenhui Yan"
date: "2024-10-13"
output: github_document
---
# Load Libraries and Set Plot Settings
```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(patchwork) # For combining plots
library(ggridges)
library(hexbin)     # For hexbin plots
library(p8105.datasets)

# Set global plot options
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = .6,
  out.width = '90%'
)

options(
  ggplot2.continuous.color = 'viridis',
  ggplot2.continuous.fill = 'viridis'
)

scale_color_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

# Problem 1
## Part1: Load and describe the Data:
```{r}
data("ny_noaa")
# View the dimensions of the dataset
dim(ny_noaa)
# View the structure of the dataset
str(ny_noaa)
# Summarize the dataset
summary(ny_noaa)
```
### A short description of the dataset:
* The ny_noaa dataset contains 2,595,176 observations and 7 variables.
* Variables:
  * id: Weather station identifier.
  * date: Date of the observation.
  * prcp: Precipitation (tenths of mm).
  * snow: Snowfall (mm).
  * snwd: Snow depth (mm).
  * tmax: Maximum temperature (tenths of degrees C).
  * tmin: Minimum temperature (tenths of degrees C).

### Missing Data:
```{r}
# Calculate missing data percentages
ny_noaa %>%
  summarise(
    prcp_missing = mean(is.na(prcp)) * 100,
    snow_missing = mean(is.na(snow)) * 100,
    snwd_missing = mean(is.na(snwd)) * 100,
    tmax_missing = mean(is.na(tmax)) * 100,
    tmin_missing = mean(is.na(tmin)) * 100
  )

```
* Observations:
  * Approximately 5.62% of prcp values are missing.
  * Approximately 14.68% of snow values are missing.
  * Approximately 22.80% of snwd values are missing.
  * A significant portion of tmax (43.69%) and tmin (46.23%) values are missing.



